NHP Analysis for Nucleotide Data
```{r}
library(readr)
library(reshape2)
library(ggplot2)
```

```{r}
#Load in the data and make the data frame columns correct
nhp_activity_data <- read.csv("~/Desktop/NHP_nucleotide_activity/nhp_activity_data.csv", stringsAsFactors = FALSE)
colnames(nhp_activity_data)[1] <- "Nucleotide"
nhp_activity_data <- t(nhp_activity_data)
colnames(nhp_activity_data) <- nhp_activity_data[1,]
nhp_activity_data <- as.data.frame(nhp_activity_data)
nhp_activity_data <- nhp_activity_data[-c(1),]
nhp_activity_data[] <- lapply(nhp_activity_data, as.numeric)
nhp_activity_data["identifier"] <- rownames(nhp_activity_data)
melted_nhp_activity_data <- melt(nhp_activity_data, id = "identifier")
colnames(melted_nhp_activity_data) <- c("identifier", "Nucleotide", "counts")
melted_nhp_activity_data["counts"] <- as.numeric(unlist(melted_nhp_activity_data["counts"])) 
```

```{r}
#merge information abotu samples with the samples themselves
meta <- read_csv("~/Desktop/NHP_nucleotide_activity/NHPmeta2.csv")
melted_nhp_activity_data <- merge(meta, melted_nhp_activity_data, by = "identifier")
nhp_activity_w_averages <- nhp_activity_data
nhp_activity_data <- merge(meta, nhp_activity_data)

#plot the mass spec values across all the different molecules
ggplot(melted_nhp_activity_data) +
        geom_point(aes(x = Nucleotide, y = counts, color = sample_type)) +
        xlab("Mass Spec Value") +
        ylab("Normalized (by Total Protein) Mass Spec Values") +
        ggtitle("Protein Counts By Cell Line") +
        theme_bw() +
	theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
	theme(plot.title=element_text(hjust=0.5)) 

```


```{r}
# plot the cell line changes in mass spec value
for (i in c(5:26)){  
  nucleotide <- colnames(nhp_activity_data)[i]
  title <- paste0("Mass Spec Values Per Cell Line: ", nucleotide)
  cnts <- nhp_activity_data[,c(1:4, i)]
  print(ggplot(cnts,aes(x=cell_line,y=nhp_activity_data[,i],col=sample_type)) + geom_point(position=position_jitterdodge(jitter.width = 0, jitter.height = 0)) + ggtitle(title) + ylab("Mass Spec Values (Normalized to Total Protein)") +
     theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(plot.title=element_text(hjust=0.5))+
    stat_summary(mapping = aes(fill=sample_type), fun = mean, position=position_dodge(0.75),
                 geom = "point", size=2, color = "black", show.legend=FALSE ))
}
```

```{r}
# plot the mass spec values for each molecule across species
for (i in c(5:26)){  
  nucleotide <- colnames(nhp_activity_data)[i]
  title <- paste0("Mass Spec Values Per Species: ", nucleotide)
  cnts <- nhp_activity_data[,c(1:4, i)]
  print(ggplot(cnts,aes(x=species,y=nhp_activity_data[,i],col=sample_type)) + geom_point(position=position_jitterdodge(jitter.width = 0, jitter.height = 0)) + ggtitle(title) + ylab("Mass Spec Values (Normalized to Total Protein)") +
     theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(plot.title=element_text(hjust=0.5))+
    stat_summary(mapping = aes(fill=sample_type), fun = mean, position=position_dodge(0.75),
                 geom = "point", size=2, color = "black", show.legend=FALSE))
}

```

```{r}
#plot the mass spec values per species with error bars and means for each condition
nhp_activity_data$species <- factor(nhp_activity_data$species, levels=c("Human", "Chimp", "Bonobo", "Rhesus"))
for (i in c(5:26)){  
  nucleotide <- colnames(nhp_activity_data)[i]
  title <- paste0("Mass Spec Values Per Species: ", nucleotide)
  cnts <- nhp_activity_data[,c(1:4, i)]
  print(ggplot(cnts, aes(x = species, y = nhp_activity_data[,i], group = sample_type, color = sample_type)) + 
      geom_point() + stat_summary(fun=mean, geom="line") + ylab("Mass Spec Values (Normalized to Total Protein)") + stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, alpha = 0.4) +
      ggtitle(title))
}
```


STATS STARTS HERE:

```{r}
#ANOVA: I was testing this. Didn't work well. Ignore for now
#library(apex)
#library(performance)
#nested_anova <- aov(Uridine ~ species/factor(cell_line), data = control_nhp)
#check_homogeneity(nested_anova)
#new_aov2 <- aov(Uridine ~ species + (1/cell_line), data=control_nhp)
#check_homogeneity(new_aov2)
#check_sphericity(new_aov2)
#posthoc <- TukeyHSD(new_aov2)
#posthoc_species_cell_line <- na.omit(posthoc$`species:cell_line`)
#ggplot(fortify(me_aug), aes(species, Uridine, color=cell_line)) +
     #stat_summary(fun.data=mean_se, geom="pointrange") +
     #stat_summary(aes(y=.fitted), fun.y=mean, geom="line")
#this doesnt do much but was still kind of interesting to see whats different its just plotting the expected slope basically for each category
```

```{r}
#separate out the data by differnet condition
human_nhp <- nhp_activity_data[nhp_activity_data$species=="Human",]
chimp_nhp <- nhp_activity_data[nhp_activity_data$species=="Chimp",]
bonobo_nhp <- nhp_activity_data[nhp_activity_data$species=="Bonobo",]
rhesus_nhp <- nhp_activity_data[nhp_activity_data$species=="Rhesus",]
control_nhp <- nhp_activity_data[nhp_activity_data$sample_type=="CTRL",]
```

SPECIES COMPARISON:


```{r}
library(lmerTest) #this lets me get the p-values
#generate the comparisons with and without the cell line when i was deciding which model was better
list <- as.factor(colnames(nhp_activity_data))
formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type")))
human_sample_type_lm <- lapply(formula, lm, data=human_nhp)
formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type + (1 | cell_line)")))
human_sample_type_lmer <- lapply(formula, lmer, data=human_nhp)
formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ species"))) d
human_species_lm <- lapply(formula, lm, data=control_nhp)
formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ species + (1 | cell_line)")))
human_species_lmer <- lapply(formula, lmer, data=control_nhp)

```

```{r}
#Done testing. decided to stick with cell line for now.
list <- as.factor(colnames(nhp_activity_data))
formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type + (1 | cell_line)")))
human_sample_type_lmer <- lapply(formula, lmer, data=human_nhp)
formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type")))
chimp_sample_type_lm <- lapply(formula, lm, data=chimp_nhp) # NOTE THIS HAS TO CHANGE BECAUSE THERE IS ONE CELL LINE SAME FOR RHESUS
formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type + (1 | cell_line)")))
bonobo_sample_type_lmer <- lapply(formula, lmer, data=bonobo_nhp)
formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type")))
rhesus_sample_type_lmer <- lapply(formula, lm, data=rhesus_nhp)
formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ species + (1 | cell_line)")))
control_species_lmer <- lapply(formula, lmer, data=control_nhp)
```

Failed to converge are because the species are too similar i think. Make note of which ones these are below:


Test the assumptions of the models. Not all of them are a great fit. Especially worried about heterskedasity (spell) so tested for variance below.
One of them is statistically significant (variance changes). However, left as is to see if its something of interest for uniformity.


SPECIES Comparsion:
```{r}
#quality control testing for different lms also tested with LRT
library(lattice)
name_mol <- colnames(nhp_activity_data[5:26])
for (i in 1: length(control_species_lmer)){
  lmer = control_species_lmer
  plot(resid(lmer[[i]]))
  title(name_mol[i])
  print(plot(lmer[[i]]))
  title(name_mol[i])
  qqnorm(residuals(lmer[[i]]))
  print(plot(lmer[[i]],
     sqrt(abs(resid(.)))~fitted(.)))
  title(name_mol[i])
  print(plot(lmer[[i]], rstudent(.) ~ hatvalues(.)))
  title(name_mol[i])
}

```

Humans sample_type comparison:
```{r}
#quality control testing for different lms also tested with LRT
library(lattice)
for (i in 1: length(human_sample_type_lmer)){
  lmer = human_sample_type_lmer
  plot(resid(lmer[[i]]))
  title(name_mol[i])
  print(plot(lmer[[i]]))
  title(name_mol[i])
  qqnorm(residuals(lmer[[i]]))
  print(plot(lmer[[i]],
     sqrt(abs(resid(.)))~fitted(.)))
  title(name_mol[i])
  print(plot(lmer[[i]], rstudent(.) ~ hatvalues(.)))
  title(name_mol[i])
}

```

BONOBO:
```{r}
for (i in 1: length(bonobo_sample_type_lmer)){
  lmer = bonobo_sample_type_lmer
  plot(resid(lmer[[i]]))
  title(name_mol[i])
  print(plot(lmer[[i]]))
  title(name_mol[i])
  qqnorm(residuals(lmer[[i]]))
  print(plot(lmer[[i]],
     sqrt(abs(resid(.)))~fitted(.)))
  title(name_mol[i])
  print(plot(lmer[[i]], rstudent(.) ~ hatvalues(.)))
  title(name_mol[i])
}
```

CHIMP:
```{r}
for (i in 1: length(chimp_sample_type_lm)){
  lmer = chimp_sample_type_lm
  plot(lmer[[i]])
}
```

RHESUS:

```{r}
for (i in 1: length(rhesus_sample_type_lmer)){
  lmer = rhesus_sample_type_lmer
  plot(lmer[[i]])
}
```
```{r}
#Now comparing the interaction effect with and without with LRT as a proxy
nhp_lfc <- as.data.frame(NA)
list <- as.factor(colnames(nhp_activity_data))

full_formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type + species + species:sample_type + (1 | cell_line)")))
reduced_formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type + species + (1 | cell_line)")))
full_interaction <- lapply(full_formula, lmer, data=nhp_activity_data)
reduced_interaction <- lapply(reduced_formula, lmer, data = nhp_activity_data)
for (i in (c(1:length(list)-5))){
  lrt_results <- lrtest(full_interaction[[i]], reduced_interaction[[i]])
}
```

```{r}
#testing if there is a difference in variance in the two conditions and across species accounting for cell type (and not)
library(car)
library(lmtest)
list <- as.factor(colnames(nhp_activity_data))
dataset_list <- c("human_nhp", "chimp_nhp", "bonobo_nhp", "rhesus_nhp")

formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type")))
chimp_var <- lapply(formula, leveneTest, data=chimp_nhp)
print(chimp_var)
rhesus_var <- lapply(formula, leveneTest, data=rhesus_nhp)
print(rhesus_var)

#formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ species")))
#control_var <- lapply(formula, leveneTest, data=control_nhp)

formula <- lapply(list[5:26], function(x) as.formula(paste0(x, "~ sample_type + (1 | cell_line)")))
var_heteroskedacity <- lapply(formula, lmer, data = human_nhp)
for (i in 1:length(var_heteroskedacity)){
  results=leveneTest(residuals(var_heteroskedacity[[i]]) ~ human_nhp$sample_type)
  print(results)
}

formula <- lapply(list[5:26], function(x) as.formula(paste0(x, "~ sample_type + (1 | cell_line)")))
var_heteroskedacity <- lapply(formula, lmer, data = bonobo_nhp)
for (i in 1:length(var_heteroskedacity)){
  results=leveneTest(residuals(var_heteroskedacity[[i]]) ~ bonobo_nhp$sample_type)
  print(results)
}

formula <- lapply(list[5:26], function(x) as.formula(paste0(x, "~ species + (1 | cell_line)")))
var_heteroskedacity <- lapply(formula, lmer, data = control_nhp)
for (i in 1:length(var_heteroskedacity)){
  results = leveneTest(residuals(var_heteroskedacity[[i]]) ~ control_nhp$species)
  print(results)
}
```

```{r}
#testing out a log transform for UTP in humans, which had significant differences in variance. See if it changes the results of the experiment
#a lot of the QQ plots look off -- but is it worth it to log transform?
list <- as.factor(colnames(nhp_activity_data))
formula <- lapply(list[5:26], function(x) as.formula(paste0("log(",x, ") ~ sample_type + (1 | cell_line)")))
human_sample_type_lmer <- lapply(formula, lmer, data=human_nhp)
summary(human_sample_type_lmer[[21]])
#p = 0.172 for KCL so isn't significant. Old one was 
formula <- lapply(list[5:26], function(x) as.formula(paste0("log(",x, ") ~ sample_type + (1 | cell_line)")))
var_heteroskedacity <- lapply(formula, lmer, data = human_nhp)
results=leveneTest(residuals(var_heteroskedacity[[21]]) ~ human_nhp$sample_type)
print(results)
```

```{r}
#no longer significant!
#lets look at the different plots!

lmer = human_sample_type_lmer
plot(resid(lmer[[21]]))
title(name_mol[21])
print(plot(lmer[[21]]))
title(name_mol[21])
qqnorm(residuals(lmer[[21]]))
print(plot(lmer[[21]],
   sqrt(abs(resid(.)))~fitted(.)))
title(name_mol[21])
print(plot(lmer[[21]], rstudent(.) ~ hatvalues(.)))
title(name_mol[21])


```


A little bit more extra analysis looking at the interaction effect graphically...
Code is very messy and was run piecemeal. May error.

```{r}
#Average across all the different cell lines and the different conditions and the different species
#missing human control?
nhp_activity_w_averages <- nhp_activity_data
rownames(nhp_activity_w_averages) <- nhp_activity_w_averages[,"identifier"]
nhp_store <- nhp_activity_w_averages[,c(1:4)]
nhp_activity_w_averages <- nhp_activity_w_averages[,c(-1,-2,-3,-4)]
nhp_activity_w_averages[] <- lapply(nhp_activity_w_averages, as.numeric)
nhp_activity_w_averages["ADRC_Control_Avg",] <- colMeans(nhp_activity_w_averages[c("ADRC.40", "ADRC.40.1", "ADRC.40.2"),])
nhp_activity_w_averages["WT126_KCl_Avg",] <-  colMeans(nhp_activity_w_averages[c("WT126.KCl.1","WT126.KCl.2"),])
nhp_activity_w_averages["WT33_Control_Avg",] <-  colMeans(nhp_activity_w_averages[c("WT33.1","WT33.2"),])
nhp_activity_w_averages["WT33_KCl_Avg",] <-  colMeans(nhp_activity_w_averages[c("WT33.KCl.1","WT33.KCl.2"),])
nhp_activity_w_averages["WT126_Control_Avg",] <-  colMeans(nhp_activity_w_averages[c("WT126.1","WT126.2" ),])
nhp_activity_w_averages["ADRC_KCl_Avg",] <-  colMeans(nhp_activity_w_averages[c("ADRC.40.KCl.2", "ADRC.40.KCl.3"),])
nhp_activity_w_averages["Human_Control_Avg",] <-  colMeans(nhp_activity_w_averages[c("WT33.1","WT33.2","WT126.1","WT126.2","ADRC.40", "ADRC.40.1", "ADRC.40.2"),])
nhp_activity_w_averages["Human_KCl_Avg",] <-  colMeans(nhp_activity_w_averages[c("WT33.KCl.1","WT33.KCl.2","WT126.KCl.1","WT126.KCl.2","ADRC.40.KCl.2", "ADRC.40.KCl.3"),])
nhp_activity_w_averages["Bono53_Control_Avg",] <- 
colMeans(nhp_activity_w_averages[c("Bono.53.1","Bono.53.2", "Bono33.3"),])
nhp_activity_w_averages["Bono53_KCl_Avg",] <-  colMeans(nhp_activity_w_averages[c("Bono.53.KCl.1","Bono.53.KCl.2","Bono.53.KCl.3"),])
nhp_activity_w_averages["Bono86_Control_Avg",] <-  colMeans(nhp_activity_w_averages[c("Bono.86.1" ,"Bono.86.2"),])
nhp_activity_w_averages["Bono86_KCl_Avg",] <-  colMeans(nhp_activity_w_averages[c("Bono.86.KCl.1","Bono.86.KCl.2","Bono.86.KCl.3"),])
nhp_activity_w_averages["Chimp_Control_Avg",] <-  colMeans(nhp_activity_w_averages[c("Chimp.1","Chimp.2","Chimp.3"),])
nhp_activity_w_averages["Chimp_KCl_Avg",] <-  colMeans(nhp_activity_w_averages[c("Chimp.KCl.1","Chimp.KCl.2","Chimp.KCl.3"),])
nhp_activity_w_averages["Rhesus_Control_Avg",] <-  colMeans(nhp_activity_w_averages[c("Rhesus.iPS.iN","Rhesus.iPS.iN.1","Rhesus.iPS.iN.2"),])
nhp_activity_w_averages["Rhesus_KCl_Avg",] <-  colMeans(nhp_activity_w_averages[c("Rhesus.iPS.iN.KCl.1","Rhesus.iPS.iN.KCl.2","Rhesus.iPS.iN.KCl.3"),])

```

```{r}
#get meta data about the differnet samples and put it into nhp_activity_only_averages
nhp_activity_only_averages <- nhp_activity_w_averages[c(-c(1:36)),]
tmp <- strsplit(rownames(nhp_activity_only_averages), split= "_")
tmp2 <- do.call(rbind, tmp)
nhp_activity_only_averages$species <- tmp2[,1]
nhp_activity_only_averages$condition <- tmp2[,2]
```


```{r}
#calculate the FC difference between KCL and Control for each species
nhp_deltas <- as.data.frame(NA)
nhp_deltas["WT126_Delta",] <- NA
nhp_deltas["WT33_Delta",] <- NA
nhp_deltas["ADRC40_Delta",] <- NA
nhp_deltas["Bono53_Delta",] <- NA
nhp_deltas["Bono86_Delta",] <- NA
nhp_deltas["Chimp_Delta",] <- NA
nhp_deltas["Rhesus_Delta",] <- NA
nhp_deltas["Human_Delta",] <- NA
nhp_deltas["Bonobo_Delta",] <- NA
nhp_deltas["Human_Delta",c(1:22)] <- nhp_activity_only_averages["Human_KCl_Avg",c(1:22)]/nhp_activity_only_averages["Human_Control_Avg",c(1:22)]
colnames(nhp_activity_only_averages)[1] <- "Adenosine"
nhp_deltas["Human_Delta",c(23:24)] <- c("Human", "Delta")
nhp_deltas["Rhesus_Delta",c(1:22)] <- nhp_activity_only_averages["Rhesus_KCl_Avg",c(1:22)]/nhp_activity_only_averages["Rhesus_Control_Avg",c(1:22)]
nhp_deltas["Rhesus_Delta",c(23:24)] <- c("Rhesus", "Delta")
nhp_deltas["Chimp_Delta",c(1:22)] <- nhp_activity_only_averages["Chimp_KCl_Avg",c(1:22)]/nhp_activity_only_averages["Chimp_Control_Avg",c(1:22)]
nhp_deltas["Chimp_Delta",c(23:24)] <- c("Chimp", "Delta")
nhp_deltas["Bonobo_Delta",c(1:22)] <- nhp_activity_only_averages["Bono_KCl_Avg",c(1:22)]/nhp_activity_only_averages["Bono_Control_Avg",c(1:22)]
nhp_deltas["Bonobo_Delta",c(23:24)] <- c("Bonobo", "Delta")
nhp_deltas <- nhp_deltas[-1,]
nhp_deltas$species <- c("Chimp", "Rhesus", "Human", "Bonobo")
nhp_deltas$species <- factor(nhp_deltas$species, levels=c("Human", "Chimp", "Bonobo", "Rhesus"))
```


```{r}
#now calcualte it for the cell lines!
nhp_FC <- as.data.frame(NA)
nhp_activity_w_averages[] <- lapply(nhp_activity_w_averages, as.numeric)
nhp_FC["WT126",] <- NA
nhp_FC["WT33",] <- NA
nhp_FC["ADRC40",] <- NA
nhp_FC["Bonobo_53",] <- NA
nhp_FC["Bonobo_86",] <- NA
nhp_FC["Chimp",] <- NA
nhp_FC["Rhesus",] <- NA
nhp_FC ["WT126",c(1:22)] <- nhp_activity_w_averages["WT126_KCl_Avg",c(1:22)]/nhp_activity_w_averages["WT126_Control_Avg",c(1:22)]

nhp_FC ["WT33",c(1:22)] <- nhp_activity_w_averages["WT33_KCl_Avg",c(1:22)]/nhp_activity_w_averages["WT33_Control_Avg",c(1:22)]

nhp_FC ["Bonobo_53",c(1:22)] <- nhp_activity_w_averages["Bono53_KCl_Avg",c(1:22)]/nhp_activity_w_averages["Bono53_Control_Avg",c(1:22)]

nhp_FC ["Bonobo_86",c(1:22)] <- nhp_activity_w_averages["Bono86_KCl_Avg",c(1:22)]/nhp_activity_w_averages["Bono86_Control_Avg",c(1:22)]

nhp_FC ["Chimp",c(1:22)] <- nhp_activity_w_averages["Chimp_KCl_Avg",c(1:22)]/nhp_activity_w_averages["Chimp_Control_Avg",c(1:22)]

nhp_FC ["Rhesus",c(1:22)] <- nhp_activity_w_averages["Rhesus_KCl_Avg",c(1:22)]/nhp_activity_w_averages["Rhesus_Control_Avg",c(1:22)]

nhp_FC ["ADRC40",c(1:22)] <- nhp_activity_w_averages["ADRC_KCl_Avg",c(1:22)]/nhp_activity_w_averages["ADRC_Control_Avg",c(1:22)]
nhp_FC <- nhp_FC[-1,-1]
nhp_FC$species <- c("Human", "Human", "Human", "Bonobo", "Bonobo", "Chimp", "Rhesus")
nhp_FC$species <- factor(nhp_FC$species, levels=c("Human", "Chimp", "Bonobo", "Rhesus"))
```

```{r}
#plot FC across species but may error
for (i in c(1:22)){  
    colnames(nhp_FC)[1] = "Adenosine"
    nucleotide=colnames(nhp_FC[i])
    print(ggplot(nhp_FC, aes(x=species, y=nhp_only_averages[,i])) +geom_point() + ggtitle(paste0("Control/KCl Means Across Species: ", nucleotide)) +ylab(paste0("Fold Change in ", nucleotide, " Amount Across Species")))
}
```

```{r}
#plot fold change across species/cell_line
for (i in c(1:22)){  
    nucleotide=colnames(nhp_FC[i])
    print(ggplot(nhp_FC, aes(x=species,y=nhp_FC[,i])) + geom_point() + ggtitle(paste0("KCl:Control Fold Change Across Species: ", nucleotide))+ylab(paste0("Fold Change in ", nucleotide)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(plot.title=element_text(hjust=0.5))+
    stat_summary(fun = mean, position=position_dodge(0.75),geom = "point", size=2, color = "blue", show.legend=FALSE))
}
```

Extra Analysis After meeting with Jeff:
1. xanthosine: do a paired t-test
2. all monophosphates on a graph
3. all A*s (ATP, AMP, dATP etc) on a graph
4. Human vs Nonhuman change in species/kcl for CMP

```{r}
#Check if xanthosine is significantly different with KCl vs Not in Humans w/ a t-test
#averaged across samples to do pairing (so have one average for each technical replicate):
list_control <- c(nhp_activity_only_averages["ADRC_Control_Avg", "Xanthosine"], nhp_activity_only_averages["WT33_Control_Avg", "Xanthosine"], nhp_activity_only_averages["WT126_Control_Avg", "Xanthosine"])
list_kcl <- c(nhp_activity_only_averages["ADRC_KCl_Avg", "Xanthosine"], nhp_activity_only_averages["WT33_KCl_Avg", "Xanthosine"], nhp_activity_only_averages["WT126_KCl_Avg", "Xanthosine"])
#nested anova
anova <- aov(human_nhp$Xanthosine ~ human_nhp$sample_type)
summary(anova) #unbalanced
#Shapiro-Wilks Test:
shapiro.test(list_control)
shapiro.test(list_kcl)
#should i be doing them separately? regardless, appear normal. no log correction
#Paired t-test:
print(t.test(list_control, list_kcl, paired = TRUE, alternative = "two.sided"))
```

```{r}
list_monops <- c("CMP", "AMP", "GMP_75", "TMP_75", "UMP_75")
nhp_monops <- nhp_activity_data[,c("identifier", list_monops)]
melted_nhp_monops <- melt(nhp_monops, id = "identifier")
colnames(melted_nhp_monops) <- c("identifier", "monop", "counts")
melted_nhp_monops <- merge(melted_nhp_monops, meta)
melted_nhp_monops_norm <- melted_nhp_monops
melted_nhp_monops_norm <- (melted_nhp_monops_norm %>% group_by(monop) %>% mutate(Norm_Count = counts/mean(counts))) 
melted_nhp_monops$species <- factor(melted_nhp_monops$species, levels=c("Human", "Chimp", "Bonobo", "Rhesus"))
melted_nhp_monops_norm$species <- factor(melted_nhp_monops$species, levels=c("Human", "Chimp", "Bonobo", "Rhesus"))
title <- "Monophosphate Mass Spec Values Across Species"
  print(ggplot(melted_nhp_monops, aes(x = species, y = counts, group = monop, color = monop)) + 
      geom_point() + stat_summary(fun=mean, geom="line") + ylab("Mass Spec Values (Normalized to Total Protein)") + stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, alpha = 0.4) +
      ggtitle(title)) + facet_wrap(~sample_type)
title <- "Monophosphate Mass Spec Values Across Species (Normalized)"
   print(ggplot(melted_nhp_monops_norm, aes(x = species, y = Norm_Count, group = monop, color = monop)) + 
      geom_point() + stat_summary(fun=mean, geom="line") + ylab("Values Normalized to Monophosphate-Specific Mean") + stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, alpha = 0.4) +
      ggtitle(title)) + facet_wrap(~sample_type)
```

```{r}
list_monops <- c("AMP", "ATP", "ADP", "Adenosine")
nhp_monops <- nhp_activity_data[,c("identifier", list_monops)]
melted_nhp_monops <- melt(nhp_monops, id = "identifier")
colnames(melted_nhp_monops) <- c("identifier", "monop", "counts")
melted_nhp_monops <- merge(melted_nhp_monops, meta)
melted_nhp_monops_norm <- melted_nhp_monops
melted_nhp_monops_norm <- (melted_nhp_monops_norm %>% group_by(monop) %>% mutate(Norm_Count = counts/mean(counts))) 
melted_nhp_monops$species <- factor(melted_nhp_monops$species, levels=c("Human", "Chimp", "Bonobo", "Rhesus"))
melted_nhp_monops_norm$species <- factor(melted_nhp_monops$species, levels=c("Human", "Chimp", "Bonobo", "Rhesus"))
title <- "Mass Spec Values Across Species"
  print(ggplot(melted_nhp_monops, aes(x = species, y = counts, group = monop, color = monop)) + 
      geom_point() + stat_summary(fun=mean, geom="line") + ylab("Mass Spec Values (Normalized to Total Protein)") + stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, alpha = 0.4) +
      ggtitle(title)) + facet_wrap(~sample_type)
title <- "Mass Spec Values Across Species (Normalized)"
   print(ggplot(melted_nhp_monops_norm, aes(x = species, y = Norm_Count, group = monop, color = monop)) + 
      geom_point() + stat_summary(fun=mean, geom="line") + ylab("Values Normalized to Molecule-Specific Mean") + stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, alpha = 0.4) +
      ggtitle(title)) + facet_wrap(~sample_type)
```

```{r}
list <- as.factor(colnames(nhp_activity_data))
human_not_data <- nhp_activity_data
levels(human_not_data$species) <- c(levels(human_not_data$species), "NHP") 
human_not_data$species[human_not_data$species=="Bonobo"] = factor("NHP")
human_not_data$species[human_not_data$species=="Chimp"] = factor("NHP")
human_not_data$species[human_not_data$species=="Rhesus"] = factor("NHP")
human_not_data_control <- human_not_data[human_not_data$sample_type == "CTRL",]

#check if the interaction effect is significant with two species
full_formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type + species + species:sample_type + (1 | cell_line)")))
reduced_formula <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ sample_type + species + (1 | cell_line")))
full_interaction <- lapply(full_formula, lmer, data=human_not_data)
reduced_interaction <- lapply(reduced_formula, lm, data = human_not_data)
#CMP only to see if interaction is better:
lrt_results <- lrtest(full_interaction[[7]], reduced_interaction[[7]])
summary(full_interaction[[7]])
#it is not better

#check if there is a significant species effect for CMP with just two species categories
species_effect <- lapply(list[5:26], function(x) as.formula(paste0(x, " ~ species + (1 | cell_line)")))
species_res <- lapply(species_effect, lmer, data=human_not_data_control)
summary(species_res[[7]])
#there is not
```